Hydras \& Co.: Formalized mathematics in Coq\\ for inspiration and entertainment

### 2021-12-21 doc/JFLA2022/Slides/Slides.tex
JFLA Slides (draft)

\definecolor{plugincolor}{rgb}{0.6,0.0,0.6}
\definecolor{mathcolor}{rgb}{0.0,0.0,0.6}
\definecolor{coqstylecolor}{rgb}{0.0,0.0,01.0}
\definecolor{lightblue}{rgb}{0.2,0.2,1.0}
\definecolor{metavarcolor}{rgb}{0.5,0.0,1.0}
\definecolor{darkgreen}{rgb}{0.1,0.7,0.1}
\definecolor{answercolor}{rgb}{.08,.15,.8}
\definecolor{normalcolor}{rgb}{0.0,0.0,0.0}
\definecolor{exbluecolor}{rgb}{0.1,0.1,0.9}
\definecolor{dontlookcolor}{rgb}{0.5,0.5,0.5}
\definecolor{termcolor}{rgb}{0.0,0.1,0.9}
\definecolor{lookcolor}{rgb}{0.9,0.1,0.0}
\definecolor{prooftermcolor}{rgb}{0.3,0.1,1.0}
\definecolor{typecolor}{rgb}{1.0,0.6,0.0}
\definecolor{taccolor}{rgb}{0.70,0.10,0.0}
\definecolor{pink}{rgb}{0.8,0.6,0.6}
\definecolor{darkmagenta}{rgb}{0.4,0.0,0.6}
\definecolor{darkblue}{rgb}{0.0,0.0,0.6}

@@

\usepackage{ifpdf}
\ifpdf
\usepackage{graphicx}
%\usepackage{aeguill}
\else
\usepackage[dvips]{graphicx}
\fi

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%% For Alectryon

\usepackage{texments}
%%% for movies by alectryon
\usepackage{./alectryon}
\usepackage{../pygments}

@@

\newcommand{\pathtomovies}{.}%/movies}

@@

% End of Alectryon stuff
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\newcommand{\canonseq}[2]{\mbox{$\{#1\}(#2)$}}
\title{Hydras \& Co.: Formalized mathematics in Coq\\
 for inspiration and entertainment
}
\date{JFLA, February 2022}
\author{
Pierre Castéran \inst{1}
\and
    Jérémy Damour \inst{2}
\and
Karl Palmskog \inst{3}
\and Clément Pit-Claudel \inst{4}
\and Théo Zimmermann \inst{5}
}

@@

\usepackage{draftwatermark} 
\setbeamercolor{background canvas}{bg=}%transparent canvas

\begin{document}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame}
  \maketitle
\end{frame}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame}
  \frametitle{Coq-community}
  \begin{block}{Objectives}
    \begin{itemize}
    \item \TODO{}{}
    \item\TODO{}{}
      \item Documentation on libraries
    \end{itemize}
  \end{block}
\end{frame}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame}
  \frametitle{Yet another book on \coq?}
  \begin{block}{Kinds of books}
    \TODO{}{}
  \end{block}
\end{frame}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame}
  \frametitle{Example-driven books (1)}
  \begin{block}{}
    \begin{itemize}
    \item \TODO{}{Definition}
    \item No need to be complete (there is already a reference manual)
      \item No need to introduce \coq (there are already a lot of initiation books and tutorials)
    \item \coq patterns (formalization and proof) are presented  when they are needed.
    \item Allow to compare several different formalizations of a given notion, and several proofs of the same theorem.
    \item By definition, the authors choose topics they like to teach.
      \item \TODO{}{topics adapted to the potential readers?}
    \end{itemize}
  \end{block}
\end{frame}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% \begin{frame}
%   \frametitle{Example-driven books (2)}
%   \begin{block}{Tsukuba Coq User's Group}
%     Regular languages, lambda-calculus, programming language semantics, \dots

%    \includegraphics[height=35mm]{tsukubabooks.jpg}
%    %\includegraphics[height=45mm]{leaflet.jpg}
%      \end{block}
%      \begin{block}{\Hydras}
%        Kirby \& Paris'  hydra battles, exponentiation algorithms, \dots
%      \end{block}

% \end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame}
  \frametitle{What hydras are talking about?}
  \begin{block}{}
    \Hydras presents a consistent set of examples, which allow us to formalize 
    some \textcolor{mathcolor}{discrete math}, discuss
    \textcolor{coqstylecolor}{\coq style, specification and proof patterns}, and use\footnote{and depend on!} \textcolor{plugincolor}{a few plug-ins}.
  \end{block}
  \begin{block}{}
    \begin{description}
    \item[hydra battles:]

@@

 \item[addition chains (exponentiation algorithms):]
      \textcolor{mathcolor}{algorithmics},
       \textcolor{mathcolor}{arithmetic},
            \textcolor{coqstylecolor}{operational type classes},
    \textcolor{coqstylecolor}{monadic notations},
    \textcolor{coqstylecolor}{PHOAS},
  \textcolor{coqstylecolor}{parametricity}, 
        \textcolor{plugincolor}{paramcoq}
        \end{description}
  \end{block}
\end{frame}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Example: Kirby \& Paris' hydras}
\begin{frame}
  \frametitle{Example: Kirby \& Paris' hydras}
 %  \begin{block}{}

%   \begin{enumerate}
%   \item Intuitive presentation
%     \item From discrete math to \coq
%   \end{enumerate}
%   \end{block}

% \end{frame}
% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% \begin{frame}
%   \frametitle{The life of $\omega^{\omega^2+1}+1$}

  \begin{figure}[h]
  \centering
  \begin{tikzpicture}[very thick, scale=0.25]
  \node (h1) at (4,0){$\bullet$};
  \node[blue] (h2) at (4,2){$\bullet$};
  \node[blue] (h3) at (2,4){$\bullet$};
  \node[blue] (h4) at (0,6){$\bullet$};
  \node[red!90!black] (h7) at (6,2){$\bullet$};
  \node[blue] (h5) at (4,6){$\bullet$};
  \node[blue] (h6) at (6,4){$\bullet$};
  \draw (h1) -- (h2) ;
  \draw (h2) -- (h3) ;
  \draw (h1) -- node[red!90!black,font=\small,sloped,shift={(0.01,-0.075)},rotate=90]{\textbf{\ding{34}}} (h7);
  \draw (h3) -- (h4) ;
  \draw (h3) -- (h5) ;
  \draw (h2) -- (h6);

  \node (h11) at (14,0){$\bullet$};
  \node[blue] (h12) at (14,2){$\bullet$};
  \node[blue] (h13) at (12,4){$\bullet$};
  \node[blue] (h14) at (10,6){$\bullet$};
  \node[blue] (h15) at (14,6){$\bullet$};
  \node[red!90!black] (h16) at (16,4){$\bullet$};
  \draw (h11) -- (h12) ;
  \draw (h12) -- (h13) ;
  \draw (h12) -- node[red!90!black,font=\small,sloped,shift={(0.01,-0.075)},rotate=90]{\textbf{\ding{34}}} (h16);
  \draw (h13) -- (h14) ;
  \draw (h13) -- (h15) ;

\node (hn1) at (24,0){$\bullet$};
\node[blue] (hn2) at (22,2) {$\bullet$};
\node[blue] (hn3) at (22,4) {$\bullet$};
\node[blue] (hn4) at (20,6){$\bullet$};
  \node[blue] (hn5) at (22,6){$\bullet$};
  \draw (hn1) -- (hn2) ;
  \draw (hn2) -- (hn3) ;
  \draw (hn3) -- (hn4) ;
  \draw (hn3) -- (hn5) ;
\node [blue](hn2b) at (24,2) {$\bullet$};
\node[blue] (hn3b) at (24,4) {$\bullet$};
\node [blue](hn4b) at (23.4,6){$\bullet$};
  \node [blue](hn5b) at (24.6,6){$\bullet$};
  \draw (hn1) -- (hn2b) ;
  \draw (hn2b) -- (hn3b) ;
  \draw (hn3b) -- (hn4b) ;
  \draw (hn3b) -- (hn5b) ;
  \node [blue](hn2c) at (26,2) {$\bullet$};
\node [blue](hn3c) at (26,4) {$\bullet$};
\node [blue](hn4c) at (26,6){$\bullet$};
  \node [red!90!black] (hn5c) at (28,6){$\bullet$};
  \draw (hn1) -- (hn2c) ;
  \draw (hn2c) -- (hn3c) ;
  \draw (hn3c) -- (hn4c) ;
  \draw (hn3c) -- node[red!90!black,font=\small,sloped,shift={(0.01,-0.075)},rotate=90]{\textbf{\ding{34}}} (hn5c) ;
\end{tikzpicture}

@@

\begin{figure}[h]
  \centering

\begin{tikzpicture}[very thick, scale=0.25]
\node (hn1) at (4,0){$\bullet$};
\node[blue] (hn2) at (2,2) {$\bullet$};
\node[blue] (hn3) at (2,4) {$\bullet$};
\node[blue] (hn4) at (0,6){$\bullet$};
  \node[blue] (hn5) at (2,6){$\bullet$};
  \draw (hn1) -- (hn2) ;
  \draw (hn2) -- (hn3) ;
  \draw (hn3) -- (hn4) ;
  \draw (hn3) -- (hn5) ;
\node [blue](hn2b) at (4,2) {$\bullet$};
\node [blue](hn3b) at (4,4) {$\bullet$};
\node [blue](hn4b) at (3.4,6){$\bullet$};
  \node [blue] (hn5b) at (4.6,6){$\bullet$};
  \draw (hn1) -- (hn2b) ;
  \draw (hn2b) -- (hn3b) ;
  \draw (hn3b) -- (hn4b) ;
  \draw (hn3b) -- (hn5b) ;
  \node [blue](hn2c) at (6,2) {$\bullet$};
\node [blue](hn3c) at (6,4) {$\bullet$};
\node [blue](hn4c) at (6,6){$\bullet$};
 % \node (hn5c) at (8,6){$\bullet$};
  \draw (hn1) -- (hn2c) ;
  \draw (hn2c) -- (hn3c) ;
  \draw (hn3c) -- (hn4c) ;
  % \draw (hn3c) -- (hn5c) ;
  \node [blue](hn3d) at (8,4) {$\bullet$};
  \draw (hn2c) -- (hn3d) ;
  \node [blue](hn4d) at (8,6) {$\bullet$};
  \draw (hn3d) -- (hn4d) ;

@@

  \node [blue](hn3f) at (12,4) {$\bullet$};
  \draw (hn2c) -- (hn3f) ;
  \node [red!90!black](hn4f) at (12,6) {$\bullet$};
  \draw (hn3f) -- node[red!90!black,font=\small,sloped,shift={(0.01,-0.075)},rotate=90]{\textbf{\ding{34}}} (hn4f) ;

@@

 \node [black](hn4h) at (21,0) {$\bullet$};
\node [red!90!black](hn4g) at (21,2) {$\bullet$};
   \draw (hn4h) -- node[red!90!black,font=\small,sloped,shift={(0.01,-0.075)},rotate=90]{\textbf{\ding{34}}} (hn4g) ;

  \node [black](hn4f) at (24,0) {$\bullet$};
\end{tikzpicture}
\caption{At $t=3,\dots,\textcolor{red}{x}-1,\textcolor{red}{x=?}$
\label{life2}}
\end{figure}
\end{frame}

\begin{frame}
  \frametitle{Mathematical formalization of hydra battles}
   \begin{block}{}

    \begin{itemize}
    \item   We associate a hydra to each ordinal \textcolor{mathcolor}{$\alpha<\epsilon_0$} (in Cantor normal form). For instance, the first hydra of our example is the image of \textcolor{mathcolor}{$omega^{\omega^2+1}+1$}.
    \item A round (at time $t$) of a battle transforms an hydra associated with \textcolor{mathcolor}{$\alpha$} into the hydra associated with \textcolor{mathcolor}{$\canonseq{\alpha}{t+1}$}, the $t+1$-st element of the \emph{canonical sequence} of \textcolor{mathcolor}{$\alpha$}.
    \item We want to ``compute'' the length of sequences of the form 
\textcolor{mathcolor}{$\langle \alpha_k=\alpha,\dots,\alpha_{i+1}=\canonseq{\alpha_i}{i+1},\dots, 1,0\rangle$}, where \textcolor{mathcolor}{$k\in\mathbb{N}$}.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame}

 \begin{block}{}
    Ordinals less than $\epsilon_0$ are represented as \emph{ordinal notations} which allow the user to write proofs by computation (comparison, canonical sequences, etc.)
    \begin{footnotesize}
      \inputsnippets{../../movies/snippets/ON_Generic/ONDef}

@@

\begin{frame}
  \begin{block}{}
    \begin{itemize}
    \item We borrow the systematic study of the sequencesfrom Ketonen \& Solovay's article
      \emph{Rapidly Growing {R}amsey Functions (1981)}.

   %    \begin{align}
%   H'_0(i) & = i\\
%   H'_\alpha(i) &= H'_{(\canonseq{\alpha}{i+1})}(i)  \quad\textit{if $\alpha$ is a limit ordinal}\\
%   H'_{\alpha}(i) &=H'_\beta(i+1) \quad\textit{if $\alpha=\beta+1$}
% \end{align}
\vspace{5pt}
      \begin{footnotesize}
        \inputsnippets{../Hprime_HprimeDef}   
      \end{footnotesize}

>>>>

\definecolor{plugincolor}{rgb}{0.6,0.0,0.6}
\definecolor{mathcolor}{rgb}{0.0,0.0,0.6}
\definecolor{coqstylecolor}{rgb}{0.0,0.0,01.0}
\definecolor{lightblue}{rgb}{0.2,0.2,1.0}
\definecolor{metavarcolor}{rgb}{0.5,0.0,1.0}
\definecolor{darkgreen}{rgb}{0.1,0.7,0.1}
\definecolor{answercolor}{rgb}{.08,.15,.8}
\definecolor{normalcolor}{rgb}{0.0,0.0,0.0}
\definecolor{exbluecolor}{rgb}{0.1,0.1,0.9}
\definecolor{dontlookcolor}{rgb}{0.5,0.5,0.5}
\definecolor{termcolor}{rgb}{0.0,0.1,0.9}
\definecolor{lookcolor}{rgb}{0.5,0.1,0.0}
\definecolor{prooftermcolor}{rgb}{0.3,0.1,1.0}
\definecolor{typecolor}{rgb}{1.0,0.6,0.0}
\definecolor{taccolor}{rgb}{0.70,0.10,0.0}
\definecolor{pink}{rgb}{0.8,0.6,0.6}
\definecolor{darkmagenta}{rgb}{0.4,0.0,0.6}
\definecolor{darkblue}{rgb}{0.0,0.0,0.6}

@@

\usepackage{ifpdf}
\ifpdf
\usepackage{graphicx}
\else
\usepackage[dvips]{graphicx}
\fi

%%%% For Alectryon

\usepackage{texments}
\usepackage{./alectryon}
\usepackage{../pygments}

@@

\newcommand{\pathtomovies}{.}

@@

% End of Alectryon stuff
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\newcommand{\canonseq}[2]{\mbox{$\{#1\}(#2)$}}

\title{Hydras \& Co.: Formalized mathematics in Coq\\
  for inspiration and entertainment
}
\date{JFLA, February 2022}
\author{
  Pierre Castéran \inst{1}
  \and
  Jérémy Damour \inst{2}
  \and
  Karl Palmskog \inst{3}
  \and Clément Pit-Claudel \inst{4}
  \and Théo Zimmermann \inst{5}
}

@@

\usepackage[firstpage]{draftwatermark} 
\setbeamercolor{background canvas}{bg=}%transparent canvas

\begin{document}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{frame}
  \maketitle
\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Introduction}

\begin{frame}
  \frametitle{Coq-community}
  \begin{block}{Objectives}
    \begin{itemize}
    \item \TODO{}{}
    \item\TODO{}{}
    \end{itemize}
  \end{block}
\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{frame}
  \frametitle{Yet another book on \coq?}
  \begin{block}{Kinds of books (existing)}
    \TODO{}{}
  \end{block}
\end{frame}

\begin{frame}
  \frametitle{Example-driven books}
  \begin{block}{}
    \begin{itemize}
    \item The structure is determined by the presented topics (discrete math, algorithmics, etc.). \coq technology (formalization and proof patterns) is presented when needed.
    \item No need to be complete (there is already a reference manual)
    \item No need to introduce \coq (there are already a lot of initiation books and tutorials)
    \item Allow to compare several different formalizations of a given notion, and several proofs of the same theorem.
    \end{itemize}
  \end{block}
  \begin{block}{Examples}
    \begin{itemize}
    \item Tsukuba Coq Users's Group: regular languages, $\lambda$-calculus, arithmetic, ... (in Japanese + ssreflect)
    \item Software Foundations (B. Pierce \emph{et al.})
    \end{itemize}
  \end{block}
\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{frame}
  \frametitle{What hydras are talking about?}
  \begin{block}{}
    \Hydras presents a consistent set of examples, which allow us to formalize 
    some \textcolor{mathcolor}{discrete math}, discuss
    \textcolor{coqstylecolor}{\coq style, specification and proof patterns}, and use\footnote{and depend on!} \textcolor{plugincolor}{a few plug-ins}.
  \end{block}
  \begin{block}{}
    \begin{description}
    \item[hydra battles:]

@@

    \item[addition chains (exponentiation algorithms):]
      \textcolor{mathcolor}{algorithmics},
      \textcolor{mathcolor}{arithmetic},
      \textcolor{coqstylecolor}{operational type classes},
      \textcolor{coqstylecolor}{monadic notations},
      \textcolor{coqstylecolor}{PHOAS},
      \textcolor{coqstylecolor}{parametricity}, 
      \textcolor{plugincolor}{paramcoq}
    \end{description}
  \end{block}
\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Example: Kirby \& Paris' hydras}

\begin{frame}
  \frametitle{Example: Kirby \& Paris' hydras}
  \begin{figure}[h]
    \centering
    \begin{tikzpicture}[very thick, scale=0.25]
      \node (h1) at (4,0){$\bullet$};
      \node[blue] (h2) at (4,2){$\bullet$};
      \node[blue] (h3) at (2,4){$\bullet$};
      \node[blue] (h4) at (0,6){$\bullet$};
      \node[red!90!black] (h7) at (6,2){$\bullet$};
      \node[blue] (h5) at (4,6){$\bullet$};
      \node[blue] (h6) at (6,4){$\bullet$};
      \draw (h1) -- (h2) ;
      \draw (h2) -- (h3) ;
      \draw (h1) -- node[red!90!black,font=\small,sloped,
      shift={(0.01,-0.075)},rotate=90]{\textbf{\ding{34}}} (h7);
      \draw (h3) -- (h4) ;
      \draw (h3) -- (h5) ;
      \draw (h2) -- (h6);

      \node (h11) at (14,0){$\bullet$};
      \node[blue] (h12) at (14,2){$\bullet$};
      \node[blue] (h13) at (12,4){$\bullet$};
      \node[blue] (h14) at (10,6){$\bullet$};
      \node[blue] (h15) at (14,6){$\bullet$};
      \node[red!90!black] (h16) at (16,4){$\bullet$};
      \draw (h11) -- (h12) ;
      \draw (h12) -- (h13) ;
      \draw (h12) -- node[red!90!black,
      font=\small,sloped,shift={(0.01,-0.075)},
      rotate=90]{\textbf{\ding{34}}} (h16);
      \draw (h13) -- (h14) ;
      \draw (h13) -- (h15) ;

      \node (hn1) at (24,0){$\bullet$};
      \node[blue] (hn2) at (22,2) {$\bullet$};
      \node[blue] (hn3) at (22,4) {$\bullet$};
      \node[blue] (hn4) at (20,6){$\bullet$};
      \node[blue] (hn5) at (22,6){$\bullet$};
      \draw (hn1) -- (hn2) ;
      \draw (hn2) -- (hn3) ;
      \draw (hn3) -- (hn4) ;
      \draw (hn3) -- (hn5) ;
      \node [blue](hn2b) at (24,2) {$\bullet$};
      \node[blue] (hn3b) at (24,4) {$\bullet$};
      \node [blue](hn4b) at (23.4,6){$\bullet$};
      \node [blue](hn5b) at (24.6,6){$\bullet$};
      \draw (hn1) -- (hn2b) ;
      \draw (hn2b) -- (hn3b) ;
      \draw (hn3b) -- (hn4b) ;
      \draw (hn3b) -- (hn5b) ;
      \node [blue](hn2c) at (26,2) {$\bullet$};
      \node [blue](hn3c) at (26,4) {$\bullet$};
      \node [blue](hn4c) at (26,6){$\bullet$};
      \node [red!90!black] (hn5c) at (28,6){$\bullet$};
      \draw (hn1) -- (hn2c) ;
      \draw (hn2c) -- (hn3c) ;
      \draw (hn3c) -- (hn4c) ;
      \draw (hn3c) -- node[red!90!black,
      font=\small,sloped,shift={(0.01,-0.075)},
      rotate=90]{\textbf{\ding{34}}} (hn5c) ;
    \end{tikzpicture}

@@

  \begin{figure}[h]
    \centering
    \begin{tikzpicture}[very thick, scale=0.25]
      \node (hn1) at (4,0){$\bullet$};
      \node[blue] (hn2) at (2,2) {$\bullet$};
      \node[blue] (hn3) at (2,4) {$\bullet$};
      \node[blue] (hn4) at (0,6){$\bullet$};
      \node[blue] (hn5) at (2,6){$\bullet$};
      \draw (hn1) -- (hn2) ;
      \draw (hn2) -- (hn3) ;
      \draw (hn3) -- (hn4) ;
      \draw (hn3) -- (hn5) ;
      \node [blue](hn2b) at (4,2) {$\bullet$};
      \node [blue](hn3b) at (4,4) {$\bullet$};
      \node [blue](hn4b) at (3.4,6){$\bullet$};
      \node [blue] (hn5b) at (4.6,6){$\bullet$};
      \draw (hn1) -- (hn2b) ;
      \draw (hn2b) -- (hn3b) ;
      \draw (hn3b) -- (hn4b) ;
      \draw (hn3b) -- (hn5b) ;
      \node [blue](hn2c) at (6,2) {$\bullet$};
      \node [blue](hn3c) at (6,4) {$\bullet$};
      \node [blue](hn4c) at (6,6){$\bullet$};
      \draw (hn1) -- (hn2c) ;
      \draw (hn2c) -- (hn3c) ;
      \draw (hn3c) -- (hn4c) ;
      \node [blue](hn3d) at (8,4) {$\bullet$};
      \draw (hn2c) -- (hn3d) ;
      \node [blue](hn4d) at (8,6) {$\bullet$};
      \draw (hn3d) -- (hn4d) ;

@@

      \node [blue](hn3f) at (12,4) {$\bullet$};
      \draw (hn2c) -- (hn3f) ;
      \node [red!90!black](hn4f) at (12,6) {$\bullet$};
      \draw (hn3f) -- node[red!90!black,
      font=\small,sloped,shift={(0.01,-0.075)},
      rotate=90]{\textbf{\ding{34}}} (hn4f) ;

@@

      \node [black](hn4h) at (21,0) {$\bullet$};
      \node [red!90!black](hn4g) at (21,2) {$\bullet$};
      \draw (hn4h) -- node[red!90!black,
      font=\small,sloped,shift={(0.01,-0.075)},
      rotate=90]{\textbf{\ding{34}}} (hn4g) ;

      \node [black](hn4f) at (24,0) {$\bullet$};

    \end{tikzpicture}
    \caption{At $t=3,\dots,\textcolor{red}{x}-1,\textcolor{red}{x=?}$
      \label{life2}}
  \end{figure}
\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{frame}
    \begin{block}{Mathematical formalization}

    \begin{itemize}
    \item   We associate a hydra to each ordinal \textcolor{mathcolor}{$\alpha<\epsilon_0$} (in Cantor normal form). For instance, the first hydra of our example is the image of \textcolor{mathcolor}{$\omega^{\omega^2+1}+1$}.
    \item A round (at time $t\in\mathbb{N}$) of a battle transforms an hydra associated with \textcolor{mathcolor}{$\alpha$} into the hydra associated with \textcolor{mathcolor}{$\canonseq{\alpha}{t+1}$}, the $t+1$-st element of the \emph{canonical sequence} of \textcolor{mathcolor}{$\alpha$}.
    \item We want to ``compute'' the length of sequences of the form 
      \textcolor{mathcolor}{$\langle \alpha_k=\alpha,\dots,\alpha_{i+1}=\canonseq{\alpha_i}{i+1},\dots, 1,0\rangle$}, where \textcolor{mathcolor}{$k\in\mathbb{N}$}.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{frame}
  \begin{block}{}
    \begin{itemize}
    \item Sequences associated with hydra battles are studied
      as \emph{$\alpha$-large sets of natural numbers}  in Ketonen \& Solovay's article
      \emph{Rapidly Growing {R}amsey Functions (1981)}.

      \begin{align}
        H'_0(i) & = i\\
        H'_\alpha(i) &= H'_{(\canonseq{\alpha}{i+1})}(i)  \quad\textit{if $\alpha$ is a limit ordinal}\\
        H'_{\alpha}(i) &=H'_\beta(i+1) \quad\textit{if $\alpha=\beta+1$}
      \end{align}

    \end{itemize}
  \end{block}

  \begin{block}{}
    The \textcolor{mathcolor}{$H'_\alpha$} family,  is a variant of the  \emph{Hardy hierarchy of rapidly growing functions}. For instance, \textcolor{mathcolor}{$H'_{\omega^3+3}(0)\geq2^{2^N}$}, where \textcolor{mathcolor}{$N=2^{70}-1$}.
  \end{block}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{frame}
  \frametitle{Ordinals in \coq}

  \begin{block}{}
    Ordinals less than $\epsilon_0$ are represented as \emph{ordinal notations}: comparison, canonical sequences, partition limit/successors/zero ordinals are implemented by certified functions, allowing the user to write proofs by computation.

    \begin{footnotesize}
      \inputsnippets{./ONDef}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{frame}
  \begin{block}{}
    The functions $H'_\alpha$ are ``implemented'' with \textcolor{plugincolor}{coq-equations}.
    \vspace{5pt}
    \begin{footnotesize}
      \inputsnippets{../Hprime_HprimeDef}   
    \end{footnotesize}

  \begin{block}{}
    Using combinatorial lemmas by  [K. \& S.], we prove (by transfinite induction on \textcolor{mathcolor}{$\alpha$}):
    \begin{quote}
      \begin{itemize}
      \item \textcolor{mathcolor}{$H'_\alpha$} is strictly monotonous
      \item \textcolor{mathcolor}{$\forall i, i \leq H'_\alpha(i)$}
      \item   If \textcolor{mathcolor}{$\beta<\alpha$}, then \textcolor{mathcolor}{$H'_\alpha(i)>H'_\beta(i)$} for all but a finite number of $i$ (we say that \textcolor{mathcolor}{$H'_\alpha$}
        \emph{dominates} \textcolor{mathcolor}{$H_\beta$}).
      \end{itemize}
    \end{quote}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{frame}
  \begin{block}{Theorem}
    For any   \textcolor{mathcolor}{$\alpha\geq\omega^\omega$} the function \textcolor{mathcolor}{$H'_\alpha$} is not primitive recursive.
  \end{block}

  \begin{block}{}
    {\color{coqstylecolor}
      \begin{itemize}
      \item \emph{In order to prove this theorem, we chose to host and  maintain under \community Russel O'Connor's library \textcolor{plugincolor}{goedel}, which contains a formalization of primitive recursive functions.}

      \item    \emph{The interest of O'Connor's contribution led us to write a chapter on primitive recursive functions, with [counter-]examples and exercices.}

      \item \emph{Proving formally that a given function is not primitive recursive is a nice example of application of a complex (mutually recursive, dependently typed) induction principle. You have to take into account the set of PR functions of any arity.}

      \end{itemize}
    }
  \end{block}
\end{frame}

\begin{frame}

  \begin{scriptsize}
    \inputsnippets{./SchemePrimRecInda}
  \end{scriptsize}
\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{frame}
  \begin{block}{}
    Using  \texttt{PrimRec\_PrimRecs\_ind}, we prove the following lemma:

    \begin{quote}
      For any \textcolor{mathcolor}{$n$}, and any primitive recursive function \textcolor{mathcolor}{$f$} of  arity \textcolor{mathcolor}{$n$}, there exists some natural number \textcolor{mathcolor}{$q$} such that the following inequality holds:
      {\color{mathcolor}
        \[
          \forall x_1,\dots,x_n, 
          f(x_1,\dots,\,x_n)\leq\textrm{Ack}(q,\textrm{max}(x_1,\dots,x_n))
        \]}
    \end{quote}
  \end{block}

  \begin{block}{}

    Instantiating with \textcolor{mathcolor}{$n=1$}, we obtain:

    \begin{quote}
      If \textcolor{mathcolor}{$H'_{\omega^\omega}$} were primitive recursive, there would exists some \textcolor{mathcolor}{$q$} such that, for any \textcolor{mathcolor}{$k$}, \textcolor{mathcolor}{$H'_{\omega^\omega}(k) \leq\textrm{Ack}(q,k)$}, which is false.
    \end{quote}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% 
\section{Presenting proofs with \alectr}
\begin{frame}
  \frametitle{Presenting proofs with \alectr}
  \begin{block}{}
    \begin{itemize}
    \item The order in which \coq code; goals and answers are presented in the book is determined by the difficulty of the mathematical examples, or the assumed level in \coq of the reader, not by the structure of the libraries.
    \item
      \emph{\coq libraries contain \emph{snippet declarations} which are converted during the compilation of the book into \LaTeX\xspace blocks which are inserted into the book.}

    \item {\color{lookcolor} \emph{Note that our build system guarantees that the accompanying book is always up-to-date and consistent with the \coq code.}}
    \end{itemize}
  \end{block}
\end{frame}

\begin{frame}[fragile]
\begin{alltt}
{\color{lightblue}{(* begin snippet Ex42a:: no-out *)}}
Example Ex42: omega + 42 + omega^2 = omega^2. 
{\color{lightblue}{(* end snippet Ex42a *)}}
Proof.
  {\color{lightblue}{(* begin snippet Ex42b *)}}
  assert (HAP:= AP_phi0 2). {\color{lightblue}{(* .no-out *)}}
  elim  HAP; intros _ H0; apply H0; clear H0. 
  {\color{lightblue}{(* end snippet Ex42b *)}}
  {\color{lightblue}{(* begin snippet Ex42d *)}}
  Check AP_plus_closed. {\color{lightblue}{(* .unfold .no-goals *)}}
  {\color{lightblue}{(* end snippet Ex42d *)}}
   {\color{lightblue}{(* begin snippet Ex42c *)}}
  assert (Hlt: omega < omega^2) by 
      (rewrite omega_eqn; apply phi0_mono, finite_mono;
        auto with arith).
 {\color{lightblue}{ (* end  snippet Ex42c *)}}
\end{alltt}
\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{frame}
  \begin{small}

     \inputsnippets{../Schutte_Ex42a}

     By definition of additive principal ordinals, 
    it suffices to prove the inequality $\omega+42< \phi_0(2)$.

    \inputsnippets{../Schutte_Ex42b}
    By Lemma \textit{AP\_plus\_closed}, it suffices to prove  $\omega<\omega^2$ and $42<\omega^2$.
  \end{small}

\begin{frame}
  \begin{small}
    \inputsnippets{
      ../Schutte_Ex42c, ../Schutte_Ex42d, ../Schutte_Ex42e}
  \end{small}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{frame}

  \begin{block}{Remarks}
    \begin{itemize}
    \item We could write the book, because we are also the maintainers of the cited \coq source, and could insert the snippet instructions in the \texttt{.v} files. \textcolor{lookcolor}{But what about commenting interesting code from other people (\emph{e.g}
        \gaia) ?}
    \item The same  piece of \coq code may be used in several documents, with different styles (book, beamer), and thus
      different filters (hiding or showing subgoals, hypotheses, etc.).

    \end{itemize}
  \end{block}

  \begin{block}{To do ?}
    Define a language for
    \begin{itemize}
    \item Searching sentences in any \coq file,
    \item Applying filters in order to generate as many snippets as needed.
    \end{itemize}
  \end{block}
\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{frame}
  \TODO{}{maintenance, CI/CD,  etc.   (several slides).}
\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{frame}
  \frametitle{Conclusion}
  \begin{block}{}
    \begin{itemize}
    \item Changes in definitions and proofs already happen
      (\emph{e.g.} type class structure), without making the book's maintenance difficult.
    \item \Hydras and \community are on github. Please contribute!
      \begin{itemize}
      \item New chapters
      \item Any improvement (text and/or proofs)
      \end{itemize}
    \end{itemize}
  \end{block}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{frame}[fragile]
  \frametitle{Perspectives}
  \begin{block}{Library on ordinals}
    There exist already \coq libraries which formalize ordinal numbers.
    \begin{itemize}
    \item \gaia (José Grimm) : Formalization of Bourbaki's Elements of Mathematics. {\color{lookcolor}\emph{This library uses an ordinal representation compatible with ours. It is  written in SSreflect/Mathcomp, and is also maintained on \community}.}
    \item A formalization by Minki Cho, using a different representation.
\begin{verbatim}
Inductive Ord.t :=
| Ord.build: 
    forall (A: Type), (A -> Ord.t) -> Ord.t.
\end{verbatim}
